Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-07-22 ms/manuscript.Rmd
Make major changes to climate analysis:
- Add VPD

- Square-root transform some climate variables

- Select non-correlated climate variables and diversity metrics

- Use full-subset GAMs to analyze contribution of each climate variable to each diversity metric

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included 201 days of data for 13 epiphytic dataloggers and 13 terrestrial dataloggers ($\n = 26$).

@@

We calculated mean daily mean, minimum, and standard variation of temperature and RH for each site from the raw climate data, then grand means from the daily means. We analyzed correlation between the grand means with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). We constructed models to test for differences in response of climatic variables to elevation between epiphytic and terrestrial dataloggers. For each retained climatic variable, we constructed three linear models: one with growth habit (epiphytic vs. terrestrial) only, one with elevation only, and one with the interaction of elevation and growth habit. We compared the fit of the models using the Akaike Information Criteron (AIC) [@Akaike1973], and kept the model with the smallest AIC for each climatic variable. A single outlier site located on an exposed south-facing slope with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from analysis.

@@

We analyzed community functional trait diversity using the \function{dbFD} function in the \package{FD} R package [@Laliberte2011]. This function takes as input values any number of traits for species in a given community, plots them in a multidimensional trait space using PCA, and calculates several summary statistics. Functional richness (FRic) is equivalent to convex hull volume containing all observed trait values plotted in multidimensional trait space [@Cornwell2006], functional evenness (FEve) is the average distance to the center of all the traits, and functional dispersion (FDis) is the mean distance between traits [@Villeger2008]. We chose these three metrics because they are statistically independent and can each provide different information about the structure of community trait diversity [@Savage2012]. Functional richness does not use abundance data, but functional evenness and divergence can be weighted by abundance, so we preformed analyses for these metrics with abundance weighting to correct for the effect of rare species with unusual morphologies. Selection, transformation, and scaling of traits was performed similarly as for the PCA, except that the method used to calculate distance [@Gower1971] can accommodate missing data, so we included some species with observations missing for some traits ($\n = 126$ \si{\species} total). We calculated each metric separately for epiphytic and terrestrial communities, and compared results between the two. Under a scenario of environmental filtering (as hypothesized for epiphytic communities), we expect relatively lower values for the three metrics of functional diversity [@Savage2012]. We also investigated relationships between functional trait diversity and environment with linear models by testing for dependence of each functional trait diversity metric against elevation and daily minimum RH. We chose these independent variables because we found that overall climate tended to change with elevation, and to test if minimum RH specifically had a filtering effect on diversity along the gradient.

>>>>

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included 201 days of data for 13 epiphytic dataloggers and 13 terrestrial dataloggers ($\n = 26$). We calculated vapor pressure deficit (VPD) from relative humidity and temperature as implemented in the \package{plantecophys} R package [@Duursma2015].

@@

We calculated mean daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site from the raw climate data, then grand means from the daily means. We analyzed correlation between the grand means with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate for each climatic response variable. For visualizing the results, we selected the linear model with the lowest Akaike Information Criterion (AIC) out of the set of models including interaction effects and the effect of each independent variable alone. A single outlier site located on an exposed south-facing slope with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from analysis.

@@

We analyzed community functional trait diversity using the \function{dbFD} function in the \package{FD} R package [@Laliberte2011]. This function takes as input values any number of traits for species in a given community, plots them in a multidimensional trait space using PCA, and calculates several summary statistics. Functional richness (FRic) is equivalent to convex hull volume containing all observed trait values plotted in multidimensional trait space [@Cornwell2006], functional evenness (FEve) is the average distance to the center of all the traits, and functional dispersion (FDis) is the mean distance between traits [@Villeger2008]. We chose these three metrics because they are statistically independent and can each provide different information about the structure of community trait diversity [@Savage2012]. Functional richness does not use abundance data, but functional evenness and divergence can be weighted by abundance, so we preformed analyses for these metrics with abundance weighting to correct for the effect of rare species with unusual morphologies. Selection, transformation, and scaling of traits was performed similarly as for the PCA, except that the method used to calculate distance [@Gower1971] can accommodate missing data, so we included some species with observations missing for some traits ($\n = 126$ \si{\species} total).

Prior to testing for differences in diversity metrics, we first checked for correlation between metrics with Pearson's correlation coefficient, and retained a set of metrics with absolute correlation coefficients < 0.9. We then tested for differences between epiphytic and terrestrial communities with a paired t-test. We also conducted a more thorough analysis simultaneously testing the effects of environment and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all pairwise interactions of explanatory variables, and ranked these by corrected AIC (AICc) to find the models that best explain the data. Our dataset included `r nrow(diversity_data) - nrow(grand_mean_climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all environmental variables are related to elevation (see Results), we also constructed linear models each diversity metric by growth habit with elevation as the covariate.
